Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-06 pkg/paper/Paper/draft.Rnw
rewritten 4.1 and 4.3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@111 b679b2f7-6539-45d0-a605-5ddcb976cde9

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change
  towards negative skewness.

}

@@

\fixme{comment on results here} Overall, we find four periods during which structural breaks
happened. Two of them in the 90s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
tendency towards a lower mean inflation rate, combined with a decreasing difference between
individual inflation rates.

@@

To illustrate the proposed strategy, the case of Slovenia is considered. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successfull in realigning its economy and introduced a number of reforms leading
to a stable growth in recent years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June~2004 and later on to introduce
the euro in January~2007. The seasonally adjusted inflation series for Slovenia encompasses
the time from \Sexpr{start(x$data)} to \Sexpr{end(x$data)} and is depicted in the left
panel of Figure~\ref{fig:slovenia1}.

@@

\clearpage
This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{round3(summary(x$model[[1]])$chisq.test$p.value)} and
\Sexpr{round3(summary(x$model[[2]])$chisq.test$p.value)}, respectively.

@@

\readme{questions and challenges} The aim of this paper is to test for structural breaks in the dynamics of the inflation rates. Making use of the method developed in Section~\ref{sec:model},  the effect of the EMU upon national inflation rates can be investigated. Apart from the countries of the euro--zone, the inflation dynamics of the other European countries are also of interest. 
This issue is closely related with the question about the convergence of the national inflation rates of the EMU members towards some common mean. Many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials in parts of the sample.\readme{perhaps skip this one??} \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity in terms of average inflation and its degree of persistence.

\readme{methods, approach} To explore these issues, we make use of three different approaches. First, in Section~\ref{sec:bpresults}, the estimation technique proposed in Section~\ref{sec:model} is applied to each of the countries included in the data set introduced in Section~\ref{sec:data}. Table~\ref{tab:results1} reports the estimated breakpoints along with the segment-specific moments (mean, variance, skewness) implied by the corresponding parameter
estimates. Furthermore, entry dates for ERM/ERM~II and the euro are included as additional information. For ease of interpretation, countries with similar patterns of
change are grouped together and ordered alphabetically within each group.
If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at the 5\% level. In significant cases, the number of breakpoints was selected via the LWZ criterion and subsequently all parameters (breakpoints and segment-specific GL parameters) were estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation in the empirical scores after inclusion of the breakpoints (if any).

Section~\ref{sec:aggregate} deals with the convergence issue. Table~\ref{tab:results2} gives an overview over the data aggregated into three different country groups: the core--euro countries (Austria, Belgium, Finland, France, Germany, Luxembourg, and Netherlands), the PIIGS countries (Portugal, Ireland, Italy, Greece, and Spain), and the other countries. This is done by showing mean, variance, and skewness over time and taking possible break points in the data into account.

Finally, Figure~\ref{fig:glmean} in Section~\ref{sec:fcrisis}, depicts the inflation rates and a rolling mean estimation. This is another way of investigating into a possible convergence pattern of national inflation rates and is especially useful to depict the influence of the financial crisis.

@@

\noalign{\medskip}
\multicolumn{8}{l}{\emph{Euro--area}} \\
\hline
<<moments-table6, results=tex>>=
writeLines(unlist(lapply(22, format_moments)))
@

@@

\subsection{Breakpoint estimation results}\label{sec:bpresults}

\readme{changed to: 1) individual countries (table 1), 2) EU (table 1), 3) groups of countries (table 2), 4) crisis (fig.3)}
For Table~\ref{tab:results1}, we grouped the countries according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries had and still have 
a very low inflation rate in terms of both mean and variance and their inflation dynamics
did not change. This does make economic sense since the ECB is modeled after 
the Deutsche Bundesbank and the ECB did not -- until recently -- much diverge from the 
monetary policy of the former. Thus the EMU did have no effect since the policies in these countries were
already much in line with Germany and thus the ECB.

@@

The next group consists of three of the PIIGS  countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successfull in reducing its inflation rate, whereas Luxembourg
experienced a doubling of its former -- very low -- inflation rate with the
variance rising by a factor of almost 14 with a strong shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successfull in reducing mean inflation rates. In this group, as well
as in the previous, the EMU adjustment led to a lower mean inflation rate.

The forth subgroup consists of two of the PIIGS  countries, namely Ireland and Spain.  
There is a break in both countries in the summer of  2008  following the financial crisis and the severe 
problems concerning the real estate situation. In Ireland as well as in Spain, this  
had a strong deflationary impact, with  deflation in Ireland over an extended period of time. 
In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland,  the housing prices did not adjust so abruptly and  
in such a degree.

%This deflationary  trend, which for a time affected all countries,  
%can be seen in Figure~\ref{fig:glmean},  which depicts a rolling mean estimation,  
%where the mean is calculated using the first  moment of a fitted GL distribution.  
%There is evidence of a decline in mean inflation,  which returns to its previous 
%path roughly a year later.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and in that way also not to participate in the EMU, i.e. the United Kingdom and Sweden. Both show a clear
break in the early 90s. These can be traced back to economic crisis, the currency
crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and
the banking crisis in Sweden during the early 90s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
later part of the 90s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998. This is a result
of these countries efforts to curb in inflation by decreasing the growth of the money 
supply and indicates the time when they actually overcame the biggest
transitionally shocks on their way towards free market economies. In Poland, this took two more
years. Slovakia and Slovenia are special cases: for once, they started with lower mean
inflation values and on the other hand, their inflation dynamics  both changed roughly one year
prior to their entry into the ERM~II.

@@

In the case of the inflation rate of the euro--area, we witness three different regimes. One until 
1999 with both a low mean and variance, the next shortly after the beginning of the third and final stage of
monetary integration, accompanied with a significant increase in the mean. The overheating of the economy that
culminated in the financial crisis shows itself in a big increase in variance and a lower mean inflation, reflecting
the negative inflation rates is some euro--area countries.

@@

\clearpage
\subsection{Convergence on an aggregate level}\label{sec:aggregate}

@@

@@

The time periods are chosen according to four fundamental change periods. Namely, the early 90s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 90s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. And also the period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the two countries with the most visible effect of the financial crisis (Ireland and Spain).

@@

The other countries were successfull in decreasing all three moments from \Sexpr{yrs[1]} to \Sexpr{yrs[3]}. Still, their median
inflation rate is higher than those of the core countries. Also, the interquartile range of the mean decreased considerably. 
Contrary to the euro--zone and the PIIGS countries, variance decreased also, both in terms of the median value and the
interquartile range. The skewness decreased as well, as did the interquartile range of the skewness parameter.

@@

\clearpage
\subsection{The influence of the financial crisis}\label{sec:fcrisis}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from rolling GL fits (red)
over time for all countries, with a window size of 12 months}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

Another way to address the convergence behavior of the inflation rates and the influence of the financial crisis is a graphical representation, as in Figure~\ref{fig:glmean}. It depicts the inflation rates and a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution. The financial crises and the short deflationary phase starting  around the autumn of 2007 and lasting a few months is clearly visible. It is outlined by a strong increase in mean inflation during the boom years of 2007/2008 accompanied by an equally strong deflation following the bust. However, the level of mean inflation did not change much, as the inflation rates seem to have returned to their prior path. The inflation series of the euro--area conveys the same message. 
With regard to the financial crisis, this shows that for low frequency data, like inflation, rapid changes
that persist only for a very short period of time
are not captured well. Specifically, short-lasting changes in the mean may simply be attributed
to increased variance. For example, the mean inflation of the euro--area in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
The general trend also suggests	 that the inflation rates converged towards some common -- albeit broad -- mean level. 
There is still some visible difference in the inflation rates, but the mean estimates are moving closer, albeit in a rather broad band. So contrary to \citet{palomba}, who find that the degree of similarity in short--run inflation dynamics is still weak, we find evidence of convergence within the countries in the sample.

@@

\clearpage
\section{Data and sales periods}\label{sec:sales}

>>>>

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change
  towards negative skewness.
  %% avoid "we"-constructs
  %% "adopt" better than "assume"
}

@@

\fixme{comment on results here} Overall, we find four periods during which structural breaks
happened. Two of them in the 1990s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
tendency towards a lower mean inflation rate, combined with a decreasing difference between
individual inflation rates.

@@

To illustrate the proposed strategy, the case of Slovenia is considered. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successful in realigning its economy and introduced a number of reforms leading
to a stable growth in recent years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June~2004 and later on to introduce
the euro in January~2007. The seasonally adjusted inflation series for Slovenia encompasses
the time from \Sexpr{start(x$data)} to \Sexpr{end(x$data)} and is depicted in the left
panel of Figure~\ref{fig:slovenia1}.

@@

This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{round3(summary(x$model[[1]])$chisq.test$p.value)} and
\Sexpr{round3(summary(x$model[[2]])$chisq.test$p.value)}, respectively.

@@

To investigate the effects of the EMU on national inflation rates, the
techniques expounded in Section~\ref{sec:model} are employed to
uncover changes in the dynamics of European inflation rates. The inflation
dynamics within countries from the euro-zone (but also other European
countries) is closely related with the question about the convergence
of the national inflation rates of the EMU members towards some common mean.
Many researchers provide evidence against convergence among European countries, e.g.,
\citet{hofmann} find considerable amounts of inflation differentials and
\citet{cap} describe rather strong heterogeneity among 12~EMU countries, both
in terms of average inflation and its degree of persistence.

To explore the question of convergence, we proceed in two steps:
First, in Section~\ref{sec:bpresults}, the estimation technique proposed
in Section~\ref{sec:model} is applied to each of the countries included in
the data set (see Section~\ref{sec:data}). Second, Section~\ref{sec:aggregate}
aggregates the results from each individual country to reveal different
potential convergence patterns over time within three groups of countries:
core-euro (\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}),
PIIGS  (\Sexpr{paste(colnames(hicps)[piigs], collapse = ", ")}), and the
rest.

Finally, Section~\ref{sec:fcrisis} focuses on the question of changes
in the inflation dynamics during the 2008/2009 financial crisis.

@@

\noalign{\medskip}
\multicolumn{8}{l}{\emph{Euro area}} \\
\hline
<<moments-table6, results=tex>>=
writeLines(unlist(lapply(22, format_moments)))
@

@@

\subsection{Inflation dynamics on a national level}\label{sec:bpresults}

As a first step towards the investigation of inflation convergence, the
structural breaks in the inflation dynamics within each countries are
investigated using the methods introduced in Section~\ref{sec:model}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the
segment-specific moments (mean, variance, skewness) implied by the corresponding
parameter estimates. Furthermore, entry dates for ERM/ERM~II and the euro are
included as additional information. For ease of interpretation, countries with
similar patterns of change are grouped together and ordered alphabetically
within each group. If there is only a single segment, the sup$\mathit{LM}$ test
was non-significant at the 5\% level. In significant cases, the number of
breakpoints was selected via the LWZ criterion and subsequently all parameters
(breakpoints and segment-specific GL parameters) were estimated by ML. No
HAC~correction was employed as there was only negligible autocorrelation in the
empirical scores after inclusion of the breakpoints (if any).
In Table~\ref{tab:results1} the countries are grouped according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists of the former
DEM (Deutsche Mark) zone countries, with the exception of Austria and
Luxembourg. All these countries had and still have a very low inflation rate,
both in terms of mean and variance, and their inflation dynamics
did not change significantly. This is plausible from an economic point of view
since the European central bank (ECB) is modeled after the Deutsche Bundesbank and the ECB did not
-- until recently -- diverge much from the monetary policy of the former. Thus,
the EMU did have no effect on inflation dynamics since the policies in
these countries were already much in line with Germany and hence the ECB.

@@

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variances, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, closer to the levels of the previous two groups.

The fourth subgroup consists of two of the PIIGS countries, namely Ireland and Spain.  
There is a break in both countries in the summer of 2008 that captures the effects of
the financial crisis and the severe problems concerning the real estate situation.
In both countries, this had a strong deflationary impact, with deflation in Ireland
over an extended period of time. In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland, the housing prices did not adjust so abruptly and  
so strongly.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the timing of the biggest
transitionally shocks on the way towards free market economies. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

@@

Finally, the inflation rates for the whole euro area \fixme{Is EU euro area or European union?},
as provided by the OECD, are considered and three different regimes are found: (1)~until 
1999 with both a low mean and variance, (2)~the third and final stage of monetary integration,
accompanied with a significant increase in the mean, (3)~a period encompassing
the overheating of the economy that culminated in the financial crisis, showing a big increase
in variance and a lower mean inflation.

\fixme{some concluding remarks about the EU level bridging the current gap to the next subsection}

@@

\subsection{Convergence on an aggregate level}\label{sec:aggregate}

@@

\fixme{intro needed}

\fixme{also ``zoom out'' more, a big picture needs to be established}

@@

The time periods are chosen according to four fundamental change periods. Namely, the early 1990s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 1990s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. And also the period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the two countries with the most visible effect of the financial crisis (Ireland and Spain).

@@

The other countries were successful in decreasing all three moments from \Sexpr{yrs[1]} to \Sexpr{yrs[3]}. Still, their median
inflation rate is higher than those of the core countries. Also, the interquartile range of the mean decreased considerably. 
Contrary to the euro--zone and the PIIGS countries, variance decreased also, both in terms of the median value and the
interquartile range. The skewness decreased as well, as did the interquartile range of the skewness parameter.

@@

\subsection{The influence of the financial crisis}\label{sec:fcrisis}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from
12-month rolling GL fits (red) over time for all countries.}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

The results from the previous sections convey that the financial
crisis of 2008/9 did not have a significant effect on the inflation dynamics
in most of the European countries (except Ireland and Spain).
This may be somewhat surprising and is not enitrely correct. In fact,
there is a rather clear effect but as it is very short-lived and not
persistent it is not picked up by the structural break techniques.

More specifically, the mean inflation of the euro area \fixme{see above} from a GL fit in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
Another way to bring out this pattern in a graphical way is
Figure~\ref{fig:glmean}. It depicts the inflation rates (in gray) and a 12-month rolling mean (in red),
calculated again as the first moment of a fitted GL distribution.
The strong increase in mean inflation during the boom years of 2007/2008
accompanied by an equally strong deflation following the burst of the bubble
is clearly visible for most countries. However, the level of mean inflation
did not change persistently, as the inflation rates seem to have returned to
their prior path, conveying the same message as the inflation series of the
euro area.

As for the structural break methods adopted in the previous sections,
this means that such rapid and non-persistent effects in the mean level
are not picked up from low frequency data, like inflation rates. Instead,
they may simply be attributed to increased variance.

Apart from bringing out more details about the short-term crisis effects,
Figure~\ref{fig:glmean} also provides an alternative view on the
long-range convergence behavior of the inflation rates studied in the
previous sections. Over the 1990s and early 2000s, there is a clear
general trend towards some common -- albeit broad -- mean level. 
There is still some visible difference in the inflation rates, but the mean
estimates are moving closer in a rather broad band. So contrary to
\citet{palomba}, who find that the degree of similarity in short-run inflation
dynamics is still weak, we find evidence of convergence within the countries in
the sample.

@@

\section{Data and sales periods}\label{sec:sales}
